On The Impact of Atoms of Confusion in JavaScript Code

### 2022-07-22 3_Method.tex
Reviewed the methodology section, changing the name of the first expeirment to Repeated Measures Study and adjusting the remainder of the section for consistency. I have also attempted to emphasize that the two studies were designed and conducted independently, by different groups of researchers.

By answering these research questions, we can either generalize or refute the findings about atoms of confusion already discussed in the literature (goal of research questions (RQ1) and (RQ2)). In addition, answering the third question allows us to enrich existing catalogs about atoms of confusion and discuss how often they occur in practice\castor{This should probably be removed.}. Answering these research questions also lays the foundations for the implementation of tools that can automatically transform code into cleaner versions.
%---though we postpone these results to future research work.
We conduct a mixed-methods study to answer these questions. It includes two independently designed and conducted experiments\castor{I would rather use the term ``experiment''.}: a replication of the first study on atoms of confusion, but targeting JavaScript programs, and a new study using a latin square counterbalancing experimental design to control for subject experience. 
These two experiments explore the impact of atom candidates on understanding JavaScript code (goal of research question RQ1). They investigate whether or not programs that contain atom candidates produce a higher misunderstanding rate when programmers try to predict their outcome.\castor{I've removed the part about time because, in order for the first study to be a replication, we need to compute time differently, accounting for incorrect responses as well.}% and (b) require more time for programmers to predict their output. 
We leverage two independently-developed studies with different designs to more rigorously verify the impact of atoms of confusion in code comprehension activities. 
We also conduct a set of interviews with developers to contrast the quantitative results of the experiments with their preferences and opinions (thus addressing RQ2).

\subsection{Replication Study Setting}\label{sec:meth:replication}

We call the first study we perform the \textit{``Replication Study''} because it is a partial replication~\cite{Juristo2012} of the first study on atoms of confusion, conducted by Gopstein et. al~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17}, but targeting JavaScript instead of the C language. Gopstein et. al. employ a randomized, partial counterbalancing experimental design on a sample of 73 subjects, mostly students, with three or more months of experience in C or C++. The control variable consists of a tiny program containing a single atom candidate and the treatment a functionally-equivalent version that does not contain the atom. The main dependent variable is whether subjects are able to correctly determine the output of the programs. The other dependent variable is the time required to correctly vs. incorrectly determine the outputs. The study has the null hypothesis that \textit{``code from both control and treatment groups can be hand-evaluated with equal accuracy''} and the alternate hypothesis that \textit{``the existence of an atom of confusion causes more errors than other code in hand-evaluated outputs.''}. For each atom candidate, six programs were built, three versions including the atom candidate and three versions not including it. Each subject is exposed to four programs for every atom candidate, two including an atom and two not including an atom.

@@

In this study we adopt the same experimental design, control and treatment variables, dependent variable, and hypotheses. We have also built six programs for each atom candidate, three with and three without the atom. Considering the elements that can vary in a replication study, defined by Juristo and Gomez~\cite{Juristo2012}, in this study we employ the same experimental protocol and construct operationalizations. We have different population properties, experimenters, and a different site. In addition, since we focus on a different language, some atoms that exist in C cannot by construction exist in JavaScript programs. Conversely, JavaScript exhibits atom candidates that do not exist in C.

The design of the replication study considers two treatments: the presence or absence of atoms candidates within the programs. To identify atom candidates, we start out by selecting every atom candidate from Gopstein's study that can also happen in JavaScript. That excludes atoms related to pointers and preprocessor directives. Furthermore, we conduct an informal search of programming forums, solutions to Code Golf~\footnote{https://codegolf.stackexchange.com/} problems, code style guidelines, and expert knowledge to identify new atom candidates. Overall, the replication study analyzes 24 atoms, 15 from the original study and 9 that are new and specific to JavaScript. Table~\ref{tab:atoms_replication} presents the complete list.

For each atom candidate, we write six short programs, three including and three not including the atom candidate. The former are the control for the study, or the obfuscated versions of the programs. The alternative, atom-free, versions of the programs are the treatment, or clean, versions. Each subject is exposed to one obfuscated and one clean version of each atom, totalling 48 programs, and should determine their outputs. As mentioned before, we measure answer correctness and the time to answer questions correctly vs. incorrectly. We control for learning effects~\cite{Neely:1991:SPE} in three manners. First, by having multiple obfuscated and clean versions for each atom and presenting only one of each per subject. Second, by presenting the programs in a random order. Third, by presenting obfuscated and clean versions corresponding to the same atom candidate with at least 11 other programs in between, in  accordance to the original experimental protocol~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17}.

@@

The second section presents the programs, one at a time. For each one, there is a text box where the answer should be written and a ``Next'' button. If the subject leaves the text box empty and presses ``Next'', this is treated as a wrong answer. To reduce the likelihood of a subject attempting to execute the programs, the web application verifies whether the subject tries to change tabs or windows and presents a pop-up message if that is the case. Furthermore, it presents the programs as images, instead of text. Upon submitting an answer for a particular program, the subject is automatically led to a similar page, containing the next one.

@@

We posted the survey on the authors' social media platforms, on Brazilian CS departments' mailing lists, and on two programming subreddits. We explained our research purposes and asked developers to take the survey only if they have some familiarity with JavaScript. We collected 70 responses. The average survey respondent has a mean 26 months of experience in JavaScript, with a median of 12.\castor{We should not say this here unless we are able to do the same for the Latin square study. }

@@

This experiment uses a different experimental design that makes it possible to control for the experience of the subjects. For this study we employ a subset of the atom candidates used in the replication study. More specifically, we
%Since JavaScript and \clang have some constructs in common (Section~\ref{back}), we first 
select a set of 9+1 atom candidates: nine previously-validated atoms of confusion for the \clang language~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17} that also exist in JavaScript programs plus one atom candidate that is specific to the JavaScript language (Automatic Semicolon Insertion)\castor{I think the name of this atom is inconsistent between the studies}. 
Our supplementary material discusses and shows examples of the atoms we consider in our research.\castor{We should probably do this here or just refer to the table that I intend to insert for the replication study.}

@@

The design of our second study blocks two variables (subject experience and the programs) and
considers two treatments: the presence or absence of atom candidates within the programs. 
To achieve such a design goal, controlling the effect of experience and individual programs, we resort to the \textit{Latin Square Design} \cite{Hunter-Experimenters}. Using this design we create a 2 x 2 matrix in which each row represents a subject and each column indicates the set of programs. The design of each square (a replica) is such that no treatment is repeated in the same row or column. For example, considering that we have a set of 10
code samples $s_0, s_1, ..., s_{10}$, if a given subject (P1) is asked to predict the output of the code samples $s_0, s_1, ..., s_5$ that contain atom candidates, then, when answering questions about clean programs, they will only be presented with clean versions of the code samples $s_6, s_7,..., s_{10}$. Furthermore, a given subject (P2), who constitutes the second row of our example square, will be asked questions about the clean versions for $s_1, s_2, ..., s_5$, and will answer questions about obfuscated programs for $s_6, s_7,..., s_{10}$. By doing that, we guarantee that all versions of the programs are contained within each square, and that each configuration occurs only once within a square. Figure \ref{fig:latinsquare} offers a visual representation of the concept.

@@

For each of the 10 selected atom candidates, we write one short program containing the atom. As in the replication study, we call them the obfuscated versions of the programs. We also write corresponding short programs without the atom candidates and call them the clean versions of the programs. Overall, the experiment employed 20 programs, 10 obfuscated and 10 clean. In order to reduce the cognitive effort, each subject was asked to predict the output of 10 programs, five obfuscated and five clean. The order in which they are presented is randomized. By doing this, we seek to minimize the chances of subjects being aware that the current listing they are analyzing contains (or not) atoms of confusion. That is, each subject should indicate what would be the outcomes of the programs, some of them having atoms of confusion (while other programs do not). Since participants are not exposed to obfuscated and clean versions corresponding to the same atom candidate, learning effect is not possible. We measure answer correctness and the total time each participant requires to participate in the experiment.

@@

We implemented our survey as a web application and carried out a pilot, similarly to the replication study. 
%As part of this effort, we carried out an informal pilot whose main objectives were (i) to spot bugs in the application and in the data collection mechanism; (ii) to gain feedback from respondents about the user experience of the application; and (iii) to formulate an estimate about how long answering the survey would take on average.
Undergraduate students and professional colleagues took the pilot survey. Some users reported layout defects, and many reported that the landing page did not explain the survey well enough. We also spotted minor issues with our routines to create and populate the latin squares.

@@

The last section of the survey presents a sequence of ten questions, each containing a program. For each question, there is a text box where the answer should be written. There is also an ``I do not know'' button, which, when clicked, leads the subject to the next question. In our setting, ``I do not know'' is treated as a wrong answer. The programs are presented as images copied from a text editor, so as to demotivate respondents from resorting to external resources by copying and pasting the code into an interpreter. Upon submitting their answer for a particular question, the subjects are automatically led to a similar page, containing another program. Similarly to the replication study, we do not provide feedback about time and correctness to the subjects.

@@

\subsubsection*{Participant Selection} We invite the participants of the interviews using a snowballing technique. That is, starting from our network of contacts, we invite an initial set of candidates to take part in our survey. From this initial list, we ask for an indication of additional candidates. Our main selection criterion is that all participants should have been working with JavaScript in their daily professional activities. We invite a total of 17 developers, and 15 of them agree to participate.   
%We conducted the interviews for a period of two weeks.

>>>>

By answering these research questions, we can either generalize or refute the findings about atoms of confusion already discussed in the literature (goal of research questions (RQ1) and (RQ2)). 
\castor{The following pertains to the third RQ.}
In addition, answering the third question allows us to enrich existing catalogs about atoms of confusion and discuss how often they occur in practice\castor{This should probably be removed.}. Answering these research questions also lays the foundations for the implementation of tools that can automatically transform code into cleaner versions.
%---though we postpone these results to future research work.
We conduct a mixed-methods study to answer these questions. 
It includes two independently designed and conducted experiments: a repeated measures/within subjects study and one using a Latin square, counterbalancing experimental design to control for subject experience. By using two different experiments, designed by different researchers, with different designs, and conducted with different participants, we hope to rigorously verify the impact of atoms of confusion in code comprehension activities. These two experiments explore the impact of atom candidates on understanding JavaScript code (goal of research question RQ1). They investigate whether or not programs that contain atom candidates tend to produce more misunderstanding for programmers trying to predict their outcome.% We leverage two independently-developed and conducted experiments with different designs to more  
We also conduct a set of interviews with developers to contrast the quantitative results of the experiments with their preferences and opinions (thus addressing RQ2).

\subsection{Repeated Measures Study}\label{sec:meth:replication}

\castor{We must have introduced the atoms before.}
The first experiment we perform uses a repeated measures, or within-subject, design~\cite{Keselman:2001:ARM}. This experiment is inspired by the work of Gopstein et. al~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17}, but targeting JavaScript instead of the C language, and including some methodological differences. It covers all the atom candidates analyzed in the work of Gopstein et al. that can be transposed to JavaScript and adds candidates that are specific to JavaScript. Overall, we select 24 atom candidates in the repeated measures study.

@@

In this experiment, the control variable consists of a tiny program containing a single atom candidate and the treatment is a functionally-equivalent version that does not contain the atom. The main dependent variable is whether subjects are able to correctly determine the output of the programs. The other dependent variable is the time required to correctly vs. incorrectly determine the outputs, independently of the presence of atoms. We use the same null hypothesis as Gopstein et al.: \textit{``code from both control and treatment groups can be hand-evaluated with equal accuracy''} and the alternate hypothesis that \textit{``the existence of an atom of confusion causes more errors than other code in hand-evaluated outputs.''}. For each atom candidate, we built six programs, three versions including the atom candidate and three versions not including it. 
%Each subject is exposed to four programs for every atom candidate, two including an atom and two not including an atom.  We have also built six programs for each atom candidate, three with and three without the atom. Considering the elements that can vary in a replication study, defined by Juristo and Gomez~\cite{Juristo2012}, in this study we employ the same experimental protocol and construct operationalizations. We have different population properties, experimenters, and a different site. In addition, since we focus on a different language, some atoms that exist in C cannot by construction exist in JavaScript programs. Conversely, JavaScript exhibits atom candidates that do not exist in C.

The design of the repeated measures study considers two treatments: the presence or absence of atom candidates within the programs. To identify atom candidates, we start out by selecting every atom candidate from Gopstein's study that can also happen in JavaScript. That excludes atoms related to pointers and preprocessor directives. Furthermore, we conduct an informal search of programming forums, solutions to Code Golf~\footnote{https://codegolf.stackexchange.com/} problems, code style guidelines, and expert knowledge to identify new atom candidates. Overall, the repeated measures analyzes 24 atoms, 15 from the original study and 9 that are new and specific to JavaScript. Table~\ref{tab:atoms_replication} presents the complete list.

For each atom candidate, we write six short programs, three including and three not including the atom candidate. The former are the control for the study, or the \textbf{obfuscated} versions of the programs. The alternative, atom-free, versions of the programs are the treatment, or \textbf{clean}, versions. Each subject is exposed to one obfuscated and one clean version of each atom, totalling 48 programs, and should determine their outputs. As mentioned before, we measure answer correctness and the time to answer questions correctly vs. incorrectly. We control for learning effects~\cite{Neely:1991:SPE} in three manners. First, by having multiple obfuscated and clean versions for each atom and presenting only one of each per subject. Second, by presenting the programs in a random order. Third, by presenting obfuscated and clean versions corresponding to the same atom candidate with at least 11 other programs in between, in  accordance to the original experimental protocol~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17}.

@@

The second section presents the programs, one at a time. For each one, there is a text box where the answer should be written and a ``Next'' button. If the subject leaves the text box empty and presses ``Next'', this is treated as a wrong answer. To reduce the likelihood of a subject attempting to execute the programs, the web application verifies whether the subject tries to change tabs or windows and presents a pop-up message if that is the case. Furthermore, it presents the programs as images, instead of text, to demotivate respondents from resorting to external resources by copying and pasting the code into an interpreter. Upon submitting an answer for a particular program, the subject is automatically led to a similar page, containing the next one.

@@

We posted the survey on the authors' social media platforms, on Brazilian CS departments' mailing lists, and on two programming subreddits. We explained our research purposes and asked developers to take the survey only if they have some familiarity with JavaScript. We collected 70 responses. %The average survey respondent has a mean 26 months of experience in JavaScript, with a median of 12.\castor{We should not say this here unless we are able to do the same for the Latin square study. }

@@

This experiment uses a different experimental design that makes it possible to control for the experience of the subjects. It employs a subset of the atom candidates used in the repeated measures study. More specifically, we
%Since JavaScript and \clang have some constructs in common (Section~\ref{back}), we first 
select a set of 9+1 atom candidates: nine previously-validated atoms of confusion for the \clang language~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17} that also exist in JavaScript programs plus one atom candidate that is specific to the JavaScript language (Automatic Semicolon Insertion). 
%Our supplementary material discusses and shows examples of the atoms we consider in our research.\castor{We should probably do this here or just refer to the table that I intend to insert for the replication study.}

@@

The design of our second study blocks two variables (subject experience and the programs) and considers two treatments: the presence or absence of atom candidates within the programs. 
To achieve such a design goal, controlling the effect of experience and individual programs, we resort to the \textit{Latin Square Design}~\cite{Hunter-Experimenters}. Using this design we create a 2 x 2 matrix in which each row represents a subject and each column indicates the set of programs. The design of each square (a replica) is such that no treatment is repeated in the same row or column. For example, considering that we have a set of 10
code samples $s_0, s_1, ..., s_{10}$, if a given subject (P1) is asked to predict the output of the code samples $s_0, s_1, ..., s_5$ that contain atom candidates, then, when answering questions about clean programs, they will only be presented with clean versions of the code samples $s_6, s_7,..., s_{10}$. Furthermore, a given subject (P2), who constitutes the second row of our example square, will be asked questions about the clean versions for $s_1, s_2, ..., s_5$, and will answer questions about obfuscated programs for $s_6, s_7,..., s_{10}$. By doing that, we guarantee that all versions of the programs are contained within each square, and that each configuration occurs only once within a square. Figure \ref{fig:latinsquare} offers a visual representation of the concept.

@@

For each of the 10 selected atom candidates, we write one short program containing the atom. As in the repeated measures study, we call them the obfuscated versions of the programs. We also write corresponding short programs without the atom candidates and call them the clean versions of the programs. Overall, the experiment employed 20 programs, 10 obfuscated and 10 clean. In order to reduce the cognitive effort, each subject was asked to predict the output of 10 programs, five obfuscated and five clean. The order in which they are presented is randomized. By doing this, we seek to minimize the chances of subjects being aware that the current listing they are analyzing contains (or not) atoms of confusion. That is, each subject should indicate what would be the outcomes of the programs, some of them having atoms of confusion (while other programs do not). Since participants are not exposed to obfuscated and clean versions corresponding to the same atom candidate, learning effect is not possible. We measure answer correctness and the total time each participant requires to participate in the experiment.

@@

We implemented our survey as a web application and carried out a pilot. 
%As part of this effort, we carried out an informal pilot whose main objectives were (i) to spot bugs in the application and in the data collection mechanism; (ii) to gain feedback from respondents about the user experience of the application; and (iii) to formulate an estimate about how long answering the survey would take on average.
Undergraduate students and professional colleagues took the pilot survey. Some users reported layout defects, and many reported that the landing page did not explain the survey well enough. We also spotted minor issues with our routines to create and populate the latin squares.

@@

The last section of the survey presents a sequence of ten questions, each containing a program. For each question, there is a text box where the answer should be written. There is also an ``I do not know'' button, which, when clicked, leads the subject to the next question. In our setting, ``I do not know'' is treated as a wrong answer. The programs are also presented as images copied from a text editor. Upon submitting their answer for a particular question, the subjects are automatically led to a similar page, containing another program. Similarly to the repeated measures study, we do not provide feedback about time and correctness to the subjects.

@@

\subsubsection*{Participant Selection} We invite the participants of the interviews using a snowballing technique. That is, starting from our network of contacts, we invite an initial set of candidates to take part in our survey. From this initial list, we ask for an indication of additional candidates. Our main selection criterion is that all participants should have been working with JavaScript in their daily professional activities. We invited a total of 17 developers, and 15 of them agreed to participate.   
%We conducted the interviews for a period of two weeks.
